Modeling Galaxy Spectral Energy Distributions in the SDSS Spectroscopic Survey

### 2002-02-13 docs/paper/paper.tex
add plots
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@35 75dff457-8751-0410-ab1a-cd701654ce6b

%% Mark off your abstract in the ``abstract'' environment. In the manuscript
%% style, abstract will output a Received/Accepted line after the
%% title and affiliation information. No date will appear since the author
%% does not have this information. The dates will be filled in by the
%% editorial office after submission.
\begin{abstract}
Certain astronomical questions, such as the evolution of the
luminosity function of galaxies and the relationship between galaxy
luminosities and galaxy colors, require transforming inherently
redshift-dependent measurements of galaxy magnitudes into measures of
galaxy properties which are consistent over a range of redshifts.  To
address these astronomical questions, investigated in detail in
subsequent papers, we implement here the method of \citet{csabai00a}
for fitting spectral energy distributions (SEDs) to broad-band
photometric observations, in the context of the optical observations
of the Sloan Digital Sky Survey (SDSS). Linear combinations of four
spectral templates can reproduce the five SDSS magnitudes of galaxies
to the precision of the photometry. Expressed in the correct
coordinate system, the distribution of the coefficients of the
templates is planar, and in fact nearly linear. The resulting
reconstructed SEDs (the sum of the templates) can be used to recover
fixed-frame magnitudes over a range of redshifts. We show that the
results of doing so yield consistent results within the SDSS Main
Galaxy Sample and the SDSS Luminous Red Galaxy Sample, in the sense
that within each sample the trend with redshift of intrinsic colors
for similar type galaxies are small. A potential problem in the
measurement of the SDSS filter response curves is noted. We also
validate the results using galaxy spectrophotometry.  All code and
templates from this paper are publicly available and well-documented;
they are designed such that they can easily be incorporated into other
applications. The code is easily adapted to handle galaxy observations
in any AB magnitude system in the rest-frame optical regime.
\end{abstract}

@@

This task is nontrivial because SEDs of galaxies contain important
information on scales considerably smaller than the width of typical
broad-band filters. It is further complicated by the fact that SEDs of
distant galaxies are redshifts due to the expansion of the
universe. Thus, the more distant the galaxy, the further the observed
bandpass is blueshifted relative to the rest-frame spectral energy
distribution of the object observed.

In the past, accounting for these effects has in general come in the
form of ``$K$-corrections'' applied to the observed magnitudes. That
is, a function $K(z)$ is added to the standard cosmological bolometric
distance modulus $\mathrm{DM}(z)$. Sometimes a single function has
been applied regardless of the galaxy type, though more recently it
has become standard to use a discrete set of $K(z)$ functions
depending on galaxy type (based either on morphology or spectral
features). As described in \cite{oke68a} and later papers, the
function is based on the projection of an assumed galaxy SED
redshifted to $z$ onto the measured instrument response as a function
of wavelength (including the effects of the atmosphere, the
reflectivity of the mirrors, the filter transmission, and the response
of the CCD device or photographic emulsion).

@@

Let us begin with an SED space defined by $N_b$ template galaxy SEDs
$\vv{v}_i(\lambda)$ (say from \citealt{bruzual93a}), where $N_b$ is
large. In principle, this should be a complete set of galaxy SEDs if
we want to express {\it any} galaxy SED in this basis set; in
practice, this property is not necessary.  Given a set of
$\vv{v}_i(\lambda)$, one can find an orthonormal set of spectra
$\vv{b}_i(\lambda)$ spanning the space.

To be more precise, we define the dot-product in the space of galaxy
spectra to be:
\begin{equation}
\vv{v}_i \cdot \vv{v}_j =
\int_{\lambda_{\mathrm{min}}}^{\lambda_{\mathrm{max}}} d\lambda
v_i(\lambda) v_j(\lambda),
\end{equation}
where $\lambda_{\mathrm{min}}$ and $\lambda_{\mathrm{max}}$ define the
wavelength range used to define orthogonality. These limits are
defined not over the full range covered by the $\vv{v}_i(\lambda)$
spectra, but instead over a smaller range corresponding to the range
which is within the observed wavelength range over most of the
redshift range of the sample. Otherwise, orthogonality is defined in
terms of regions of the spectra which are poorly constrained, which
make interpretation of the fits inconvenient.  This choice does not
affect the results for the reconstructed SED fit quantitatively
whatsoever.\footnote{If the template spectra were identical in the
subrange chosen and only differed outside of it, this choice would
cause computational problems, but in fact the spectra are all
independent in the restricted, as well as the full, wavelength range.}
We then define the basis of the SED space such that the $N_b$ template
SEDs, $\vv{v}_i$, are linear combinations of $N_b$ basis spectra
$\vv{b}_i$ and that $\vv{b}_i \cdot \vv{b}_j = \delta_{ij}$. These
conditions do not fully specify the $\vv{b}_i$, which naturally may be
rotated arbitrarily within our $N_b$-dimensional subspace of spectral
space.

In our case the number of observed bands is less than $N_b$, so we
cannot determine an individual galaxy's position in this
$N_b$-dimensional space from its broad-band colors alone.  However, as
outlined by \citet{csabai00a}, one can find a lower dimensional space
defined by $N_t$ vectors $\vv{e}_k$ which best fits the {\it set} of
galaxies, as follows. The SED of galaxy $i$ may be reconstructed from a
linear combination of the basis spectra in this lower dimensional
space:
\begin{equation}
\label{lincomb}
f_{\mathrm{rec},i}(\lambda) d\lambda = 
\sum_k a_{i,k} \sum_j e_{k,j} b_j(\lambda) d\lambda,
\end{equation}
where $a_k$ are the components of the galaxy projected in the
low-dimensional space. From this SED it is easy to determine the
reconstructed flux
$F_{\mathrm{rec},il}$ in each survey bandpass by projecting
the model SED onto the bandpass $l$: 
\begin{equation}
F_{\mathrm{rec},il} = \int_0^\infty d\lambda f(\lambda) S(\lambda),
\end{equation}
where $S(\lambda)$ is the response of the instrument.  We define
\begin{equation}
\chi^2 = \sum_i \sum_l
\frac{(F_{\mathrm{obs},il}-F_{\mathrm{rec},il})^2}{\sigma_{il}^2},
\end{equation}
where $\sigma_{il}$ is the estimated error in the observed flux
$F_{\mathrm{obs},il}$ in bandpass $l$ for galaxy $i$. Taking
derivatives of $\chi^2$ with respect to $a_{i,k}$ and $e_{k,j}$
results in a set of equations bilinear in $a_{i,k}$ and $e_{k,j}$.
\citet{csabai00a} describe how to iteratively solve this bilinear
equation by first fixing $e_{k,j}$ and fitting for $a_{i,k}$, then
fixing $a_{i,k}$ and fitting for $e_{k,j}$, and iterating that
procedure.

@@

We define the ``total flux'' $f_k$ of each template as the flux in the
range $\lambda_{\mathrm{min}}<\lambda<\lambda_{\mathrm{max}}$; then an
estimate of the flux in this range is $F_{{t}}=\Sigma_k a_k f_k$. The
locus of points of a fixed flux in the component space $a_k$ lie on a
plane. Therefore, a natural choice for the orientation of the axes
$\vv{e}_k$ is to let $\vv{e}_0$ be perpendicular to the planes of
constant flux; in this manner, the coefficient $a_0$ in Equation
(\ref{lincomb}) is directly proportional to $F$ and is thus purely a
measure of the object's flux (in the wavelength range considered),
while the parameters $a_i/a_0$ (where $i>0$) primarily measure the
shape of the galaxy SED in the wavelength range considered. $F$ can be
trivially related to $L$, the luminosity in this range, by the inverse
square of the luminosity distance.  As it turns out, the parameters
$a_i/a_0$ tend to be distributed very approximately in an ellipsoid,
with not much curvature within the space defined by the template
SEDs. Thus, we rotate the axes $\vv{e}_i$ (for $i>0$) such that they
are aligned with the principal axes of the ellipsoid. Furthermore, we
normalize the vectors $\vv{e}_i$ such that the $a_0=F_t$ in units of
ergs cm$^{-2}$ s$^{-1}$. This choice of coordinate system is
arbitrary and does not affect our quantitative results, but it is
convenient.

@@

The version of the $K$-correction code (v1\_2) implementing the
calculations described here, along with eigentemplates and filter
curves is publicly available on the World Wide Web at {\tt
http://physics.nyu.edu/\~\ mb144/kcorrect}. In addition, a public CVS
repository can be accessed at {\tt xxx} (instructions for exporting
the appropriate version of the code from this repository are provided
on the web page noted above. The whole of the code can be used through
the Research Systems, Incorporated, IDL language; everything except
for the template-fitting also exists in stand-alone C programs (which
calls the same routines, guaranteeing consistency).

@@

The magnitudes are calibrated $AB$, meaning that the system is
designed such that:
\begin{eqnarray}
m_{AB} &=& -2.41 - 2.5 \log_{10}\left[
\frac{\int_{0}^{\infty} d\lambda \lambda f(\lambda) R(\lambda)}
{\int_{0}^{\infty} d\lambda \lambda^{-1} R(\lambda)}\right]
\cr
&=& -48.60 - 2.5 \log_{10}\left[
\frac{\int_{0}^{\infty} d\nu \nu^{-1} f(\nu) R(\nu)}
{\int_{0}^{\infty} d\nu \nu^{-1} R(\nu)}\right],
\cr
\end{eqnarray}
where $R(\lambda)$ is the fraction of photons entering the Earth's
atmosphere which are detected as a function of wavelength (a unitless
quantity). In this equation, $f(\lambda)$ is in units of ergs
cm$^{-2}$ s$^{-1}$ \AA$^{-1}$, and $f(\nu)$ is in units of ergs
cm$^{-2}$ s$^{-1}$ Hz$^{-1}$. The normalizations defined here mean
that an object with $f(\nu) = 3.631 \times 10^{-20}$ ergs cm$^{-2}$
s$^{-1}$ Hz$^{-1}$ has all its AB magnitudes equal to zero.  The
difference in the zeropoint in the two equations is simply the factor
of speed of light $c$ (expressed in \AA\ s$^{-1}$) between the
frequency and wavelength scales.

>>>>

%% Mark off your abstract in the ``abstract'' environment. In the manuscript
%% style, abstract will output a Received/Accepted line after the
%% title and affiliation information. No date will appear since the author
%% does not have this information. The dates will be filled in by the
%% editorial office after submission.
\begin{abstract}
Broad-band measurements of galaxy flux suffer from the difficulty that
they probe different regions of the rest-frame galaxy spectrum for
galaxies at different redshifts.  Certain astronomical questions, such
as the evolution of the luminosity function of galaxies, require
transforming these inherently redshift-dependent measurements of
galaxy magnitudes into redshift-independent quantities. To prepare to
address these astronomical questions, investigated in detail in
subsequent papers, we implement here the method of \citet{csabai00a}
for fitting spectral energy distributions (SEDs) to broad-band
photometric observations, in the context of the optical observations
of the Sloan Digital Sky Survey (SDSS). Linear combinations of four
spectral templates can reproduce the five SDSS magnitudes of galaxies
to the precision of the photometry. Expressed in the correct
coordinate system, the distribution of the coefficients of the
templates is planar, and in fact nearly linear. The resulting
reconstructed SEDs (the sum of the templates) can be used to recover
fixed-frame magnitudes over a range of redshifts. We show that the
results of doing so yield consistent results within the SDSS Main
Galaxy Sample and the SDSS Luminous Red Galaxy Sample, in the sense
that within each sample the trend with redshift of intrinsic colors
for similar type galaxies are small. A potential problem in the
measurement of the SDSS filter response curves is noted. We also
validate the results using galaxy spectrophotometry.  All code and
templates from this paper are publicly available and well-documented;
they are designed such that they can easily be incorporated into other
applications. The code is easily adapted to handle galaxy observations
in any AB magnitude system in the rest-frame optical regime.
\end{abstract}

@@

This task is nontrivial because SEDs of galaxies contain important
information on scales considerably smaller than the width of typical
broad-band filters. It is further complicated by the fact that SEDs of
distant galaxies are redshifted due to the expansion of the
universe. Thus, the more distant the galaxy, the further the observed
bandpass is blueshifted relative to the rest-frame spectral energy
distribution of the object observed.

In the past, people have in general accounted for these effects using
``$K$-corrections'' applied to the observed magnitudes. That is, a
function $K(z)$ is added to the standard cosmological bolometric
distance modulus $\mathrm{DM}(z)$. Sometimes a single function has
been applied regardless of the galaxy type, though more recently it
has become standard to use a discrete set of $K(z)$ functions
depending on galaxy type (based either on morphology or spectral
features). As described in \cite{oke68a} and later papers, the
function is based on the projection of an assumed galaxy SED
redshifted to $z$ onto the measured instrument response as a function
of wavelength (including the effects of the atmosphere, the
reflectivity of the mirrors, the filter transmission, and the response
of the CCD device or photographic emulsion).

@@

Let us begin with an SED space defined by $N_b$ template galaxy SEDs
$\vv{v}_i(\lambda)$ (say from \citealt{bruzual93a}), where $N_b$ is
large. In principle, this should be a complete set of galaxy SEDs if
we want to express {\it any} galaxy SED in this basis set; in
practice, this property is not necessary.  Given a set of
$\vv{v}_i(\lambda)$, one can find an orthonormal set of spectra
$\vv{b}_i(\lambda)$ spanning the space.  To be more precise, we define
the dot-product in the space of galaxy spectra to be:
\begin{equation}
\vv{v}_i \cdot \vv{v}_j =
\int_{\lambda_{\mathrm{min}}}^{\lambda_{\mathrm{max}}} d\lambda
v_i(\lambda) v_j(\lambda),
\end{equation}
where $\lambda_{\mathrm{min}}$ and $\lambda_{\mathrm{max}}$ define the
wavelength range used to define orthogonality. These limits are
defined not over the full range covered by the $\vv{v}_i(\lambda)$
spectra, but instead over a smaller range which is within observable
wavelengths over most of the redshift range of the sample. If we
defined orthogonality over the full range of $\vv{v}(\lambda)$, much
of which is unobservable, it would complicate the interpretation of
our fits.  This choice does not affect the results for the
reconstructed SED fit quantitatively whatsoever.\footnote{If the
template spectra were identical in the subrange chosen and only
differed outside of it, this choice would cause computational
problems, but in fact the spectra are all independent in the
restricted, as well as the full, wavelength range.}  We then define
the basis of the SED space such that the $N_b$ template SEDs,
$\vv{v}_i$, are linear combinations of $N_b$ basis spectra $\vv{b}_i$
and that $\vv{b}_i \cdot \vv{b}_j = \delta_{ij}$. These conditions do
not fully specify the $\vv{b}_i$, which naturally may be rotated
arbitrarily within our $N_b$-dimensional subspace of spectral space.

In our case the number of observed bands is less than $N_b$, so we
cannot determine an individual galaxy's position in this
$N_b$-dimensional space from its broad-band colors alone.  However, as
outlined by \citet{csabai00a}, one can find a lower dimensional space
defined by $N_t$ vectors $\vv{e}_k$ which best fits the {\it set} of
galaxies, as follows. The SED of galaxy $i$ may be reconstructed from a
linear combination of the basis spectra in this lower dimensional
space:
\begin{equation}
\label{lincomb}
f_{\mathrm{rec},i}(\lambda) d\lambda = 
\sum_k a_{i,k} \sum_j e_{k,j} b_j(\lambda) d\lambda,
\end{equation}
where $a_k$ are the components of the galaxy projected in the
low-dimensional space. From this SED it is easy to determine the
reconstructed flux
$F_{\mathrm{rec},il}$ in each survey bandpass by projecting
the model SED onto the bandpass $l$: 
\begin{equation}
F_{\mathrm{rec},il} = \int_0^\infty d\lambda
f_{\mathrm{rec},i}(\lambda) S_l(\lambda),
\end{equation}
where $S_l(\lambda)$ is the response of the instrument in band $l$.  We define
\begin{equation}
\chi^2 = \sum_i \sum_l
\frac{(F_{\mathrm{obs},il}-F_{\mathrm{rec},il})^2}{\sigma_{il}^2},
\end{equation}
where $\sigma_{il}$ is the estimated error in the observed flux
$F_{\mathrm{obs},il}$ in bandpass $l$ for galaxy $i$. Taking
derivatives of $\chi^2$ with respect to $a_{i,k}$ and $e_{k,j}$
results in a set of equations bilinear in $a_{i,k}$ and $e_{k,j}$.
\citet{csabai00a} describe how to iteratively solve this bilinear
equation by first fixing $e_{k,j}$ and fitting for $a_{i,k}$, then
fixing $a_{i,k}$ and fitting for $e_{k,j}$, and iterating that
procedure.

@@

We define the ``total flux'' $f_k$ of each template as the flux in the
range $\lambda_{\mathrm{min}}<\lambda<\lambda_{\mathrm{max}}$; then an
estimate of the flux in this range is $F_{{t}}=\Sigma_k a_k f_k$. The
locus of points of a fixed flux in the component space $a_k$ lie on a
plane. Therefore, a natural choice for the orientation of the axes
$\vv{e}_k$ is to let $\vv{e}_0$ be perpendicular to the planes of
constant flux; in this manner, the coefficient $a_0$ in Equation
(\ref{lincomb}) is directly proportional to $F$ and is thus purely a
measure of the object's flux (in the wavelength range considered),
while the parameters $a_i/a_0$ (where $i>0$) primarily measure the
shape of the galaxy SED in the wavelength range considered. $F$ can be
trivially related to $L$, the luminosity in this range, by the inverse
square of the luminosity distance.  As it turns out, the parameters
$a_i/a_0$ tend to be distributed very approximately in an ellipsoid,
with not much curvature within the space defined by the template
SEDs. Thus, we rotate the axes $\vv{e}_i$ (for $i>0$) such that they
are aligned with the principal axes of the ellipsoid. Furthermore, we
normalize the vectors $\vv{e}_i$ such that the zeroth component
$a_0=F_t$ is in units of ergs cm$^{-2}$ s$^{-1}$. This choice of
coordinate system is arbitrary and does not affect our quantitative
results, but it is convenient.

@@

The version of the $K$-correction code ({\tt v1\_2}) implementing the
calculations described here, along with eigentemplates and filter
curves is publicly available on the World Wide Web at {\tt
http://physics.nyu.edu/\~\ mb144/kcorrect}. In addition, a public CVS
repository can be accessed at {\tt xxx} (instructions for exporting
the appropriate version of the code from this repository are provided
on the web page noted above. The whole of the code can be used through
the Research Systems, Incorporated, IDL language; everything except
for the template-fitting also exists in stand-alone C programs (which
call the same routines, guaranteeing consistency).

@@

The magnitudes are calibrated $AB$, meaning that the system is
designed such that:
\begin{eqnarray}
m_{AB} &=& -2.41 - 2.5 \log_{10}\left[
\frac{\int_{0}^{\infty} d\lambda \lambda f(\lambda) R(\lambda)}
{\int_{0}^{\infty} d\lambda \lambda^{-1} R(\lambda)}\right]
\cr
&=& -48.60 - 2.5 \log_{10}\left[
\frac{\int_{0}^{\infty} d\nu \nu^{-1} f(\nu) R(\nu)}
{\int_{0}^{\infty} d\nu \nu^{-1} R(\nu)}\right],
\end{eqnarray}
where $R(\lambda)$ is the fraction of photons entering the Earth's
atmosphere which are detected as a function of wavelength (a unitless
quantity). In this equation, $f(\lambda)$ is in units of ergs
cm$^{-2}$ s$^{-1}$ \AA$^{-1}$, and $f(\nu)$ is in units of ergs
cm$^{-2}$ s$^{-1}$ Hz$^{-1}$. The normalizations defined here mean
that an object with $f(\nu) = 3.631 \times 10^{-20}$ ergs cm$^{-2}$
s$^{-1}$ Hz$^{-1}$ has all its AB magnitudes equal to zero.  The
difference in the zeropoint in the two equations is simply the factor
of speed of light $c$ (expressed in \AA\ s$^{-1}$) between the
frequency and wavelength scales.
